---
title: "Bayesian Hierarchical Modeling of Workplace Stress"
author: "Washington Casamen Nolasco"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    highlight-style: github
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
knitr:
  opts_knit:
    root.dir: "../"
---

## Overview

This report develops a Bayesian hierarchical analysis of workplace stress and well-being among early childhood educators. The project focuses on repeated measurements, uncertainty, and practical interpretation for behavioral data science and occupational psychology.

## Research Question

How do workplace demands, work resources, and time-varying factors relate to educator stress, and how much uncertainty remains after accounting for repeated observations within educators?

## Setup

```{r setup}
source("R/data_dictionary.R")
source("R/prepare_data.R")
source("R/model_functions.R")

library(dplyr)
library(ggplot2)
```

## Raw Data Inventory

```{r raw-inventory}
raw_inventory <- raw_file_inventory("data/raw")
raw_inventory
```

## Time Point Files

```{r timepoint-summary}
timepoint_data <- load_timepoint_data("data/raw")
timepoint_summary <- summarise_timepoints(timepoint_data)
timepoint_summary
```

## Column Inventory

```{r column-inventory}
column_inventory <- csv_column_inventory("data/raw")

column_inventory |>
  count(file, name = "columns") |>
  arrange(file)
```

## Analysis 1 - Longitudinal Structure

Before choosing a Bayesian model, we need to verify whether the same educators are observed across time points. This determines whether the model can use repeated-measure partial pooling, for example `(1 | StudyID)`, or whether the analysis should be treated as mostly cross-sectional by wave.

```{r analysis-1-id-audit}
id_audit <- audit_study_ids(timepoint_data, id_col = "StudyID")
id_audit
```

```{r analysis-1-id-overlap}
id_overlap <- build_id_overlap_matrix(timepoint_data, id_col = "StudyID")
id_overlap
```

**Interpretation guide:** The diagonal shows the number of unique educators in each wave. Off-diagonal cells show how many educators appear in both waves. If overlap is high, a longitudinal hierarchical model is appropriate. If overlap is low, we may need a simpler repeated-wave or available-case approach.

## Analysis 2 - Candidate Scale Scores

The raw files contain many item-level survey columns. For a first Bayesian model, we should prefer validated or precomputed scale scores when available, then verify their meaning in the codebook. This chunk searches for columns that look like totals, sums, means, averages, or scale scores.

```{r analysis-2-summary-columns}
summary_columns <- find_summary_columns(timepoint_data)

summary_columns |>
  arrange(time_point, column)
```

```{r analysis-2-scale-profiles}
scale_profiles <- profile_numeric_columns(timepoint_data, summary_columns)

scale_profiles |>
  arrange(time_point, column)
```

**Interpretation guide:** These columns are candidates, not final constructs. We will use the codebook to decide which score should become the primary outcome, such as perceived stress, burnout, depressive symptoms, job satisfaction, or teacher stress.

## Analysis 3 - Core Scale Panel

This kernel builds the first reduced longitudinal analysis table. The goal is not yet to fit a model, but to create a person-time panel with the most promising psychological scale scores. Because the raw files use different column names across time points, each construct is harmonized manually.

The cleaning step converts special missing-value codes such as `-999`, `-99`, and `-9` into real `NA` values. This matters because those codes would otherwise distort means, ranges, correlations, and Bayesian model estimates.

```{r analysis-3-core-scale-panel}
clean_missing_codes <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  x[x %in% c(-999, -99, -9)] <- NA_real_
  x
}

pick_col <- function(dat, candidates) {
  found <- candidates[candidates %in% names(dat)]

  if (length(found) == 0) {
    return(rep(NA_real_, nrow(dat)))
  }

  clean_missing_codes(dat[[found[1]]])
}

core_scale_panel <- bind_rows(
  data.frame(
    StudyID = as.character(timepoint_data$tp1$StudyID),
    time_point = "tp1",
    perceived_stress = pick_col(timepoint_data$tp1, c("PSIsumTp1")),
    teacher_stress = pick_col(timepoint_data$tp1, c("TSI_mean")),
    burnout_emotional_exhaustion = pick_col(timepoint_data$tp1, c("Burn_EEav")),
    depressive_symptoms = pick_col(timepoint_data$tp1, c("CESDsum_tp1")),
    anxiety = pick_col(timepoint_data$tp1, c("GA_TotalTp1")),
    self_efficacy = pick_col(timepoint_data$tp1, c("TBSE_totalav")),
    job_satisfaction_work = pick_col(timepoint_data$tp1, c("JS_my_work_SUM_tp1"))
  ),
  data.frame(
    StudyID = as.character(timepoint_data$tp2$StudyID),
    time_point = "tp2",
    perceived_stress = pick_col(timepoint_data$tp2, c("PSIsum_tp2")),
    teacher_stress = NA_real_,
    burnout_emotional_exhaustion = NA_real_,
    depressive_symptoms = pick_col(timepoint_data$tp2, c("CESDsum_tp2")),
    anxiety = pick_col(timepoint_data$tp2, c("GA_Total_tp2")),
    self_efficacy = pick_col(timepoint_data$tp2, c("gsE_sum_tp2")),
    job_satisfaction_work = NA_real_
  ),
  data.frame(
    StudyID = as.character(timepoint_data$tp3$StudyID),
    time_point = "tp3",
    perceived_stress = NA_real_,
    teacher_stress = pick_col(timepoint_data$tp3, c("TSI_mean_tp3")),
    burnout_emotional_exhaustion = pick_col(timepoint_data$tp3, c("Burn_EEav_tp3")),
    depressive_symptoms = NA_real_,
    anxiety = NA_real_,
    self_efficacy = pick_col(timepoint_data$tp3, c("TBSE_totalav_tp3")),
    job_satisfaction_work = pick_col(timepoint_data$tp3, c("js_my_work_SUM_tp3"))
  ),
  data.frame(
    StudyID = as.character(timepoint_data$tp4$StudyID),
    time_point = "tp4",
    perceived_stress = pick_col(timepoint_data$tp4, c("PSIsum_TP4")),
    teacher_stress = pick_col(timepoint_data$tp4, c("TSI_mean_tp4")),
    burnout_emotional_exhaustion = NA_real_,
    depressive_symptoms = pick_col(timepoint_data$tp4, c("CESDsum_tp4")),
    anxiety = pick_col(timepoint_data$tp4, c("GA_Total")),
    self_efficacy = pick_col(timepoint_data$tp4, c("TBSE_totalav_tp4")),
    job_satisfaction_work = pick_col(timepoint_data$tp4, c("js_my_work_SUM_tp4"))
  )
) |>
  mutate(
    time_point = factor(time_point, levels = c("tp1", "tp2", "tp3", "tp4")),
    time_index = as.integer(time_point)
  )

core_scale_panel |>
  summarise(
    rows = n(),
    educators = n_distinct(StudyID),
    across(
      c(
        perceived_stress,
        teacher_stress,
        burnout_emotional_exhaustion,
        depressive_symptoms,
        anxiety,
        self_efficacy,
        job_satisfaction_work
      ),
      ~ sum(!is.na(.x)),
      .names = "n_{.col}"
    )
  )
```

**Interpretation guide:** The counts show which constructs have enough observed values for modeling. A construct measured in several waves can support longitudinal modeling. A construct measured only once is better treated as a baseline predictor or descriptive endpoint.

## Analysis 4 - Scale Coverage by Time Point

The previous kernel showed total coverage across all waves. This kernel breaks coverage down by time point. This matters because a variable can have a high total count but still be unavailable in key waves. For longitudinal Bayesian modeling, the best outcome is usually a construct measured repeatedly across several time points.

```{r analysis-4-coverage-by-time}
scale_coverage_by_time <- core_scale_panel |>
  group_by(time_point) |>
  summarise(
    rows = n(),
    educators = n_distinct(StudyID),
    across(
      c(
        perceived_stress,
        teacher_stress,
        burnout_emotional_exhaustion,
        depressive_symptoms,
        anxiety,
        self_efficacy,
        job_satisfaction_work
      ),
      ~ sum(!is.na(.x)),
      .names = "n_{.col}"
    ),
    .groups = "drop"
  )

scale_coverage_by_time
```

```{r analysis-4-scale-distributions}
scale_distribution_summary <- core_scale_panel |>
  summarise(
    across(
      c(
        perceived_stress,
        teacher_stress,
        burnout_emotional_exhaustion,
        depressive_symptoms,
        anxiety,
        self_efficacy,
        job_satisfaction_work
      ),
      list(
        n = ~ sum(!is.na(.x)),
        mean = ~ mean(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE),
        min = ~ min(.x, na.rm = TRUE),
        max = ~ max(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )

scale_distribution_summary
```

```{r analysis-4-long-format-preview}
core_scale_long <- core_scale_panel |>
  tidyr::pivot_longer(
    cols = c(
      perceived_stress,
      teacher_stress,
      burnout_emotional_exhaustion,
      depressive_symptoms,
      anxiety,
      self_efficacy,
      job_satisfaction_work
    ),
    names_to = "construct",
    values_to = "value"
  ) |>
  filter(!is.na(value))

core_scale_long |>
  count(construct, time_point) |>
  arrange(construct, time_point)
```

**Interpretation guide:** A strong first outcome should have repeated coverage across at least three waves, a sensible numeric range, and no remaining special missing codes. If two outcomes are plausible, we will prefer the one that best matches the project's psychological question.

## Candidate Psychological Constructs

This first pass searches column names for likely stress, demand, resource, well-being, and educator identifiers. These candidates will be verified against the codebook before modeling.

```{r candidate-columns}
candidate_patterns <- c(
  "stress", "burn", "well", "demand", "resource", "support",
  "depress", "anx", "effic", "satisf", "job", "work", "teacher",
  "center", "site", "id"
)

candidates <- find_candidate_columns(column_inventory, candidate_patterns)

candidates |>
  arrange(file, column) |>
  select(file, column, missing_rate) |>
  head(80)
```

## Next Modeling Plan

The next step is to use the codebook to define:

- the primary stress or well-being outcome;
- educator identifier;
- center or site identifier, if available;
- repeated-measure time variable;
- job demand predictors;
- job resource predictors;
- demographic or professional covariates.

The first Bayesian model will be intentionally simple. Once variable definitions are verified, the project will move to a hierarchical longitudinal model.
