---
title: "Bayesian Hierarchical Modeling of Workplace Stress"
author: "Washington Casamen Nolasco"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    highlight-style: github
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
knitr:
  opts_knit:
    root.dir: "../"
---

## Overview

This report develops a Bayesian hierarchical analysis of workplace stress and well-being among early childhood educators. The project focuses on repeated measurements, uncertainty, and practical interpretation for behavioral data science and occupational psychology.

## Research Question

How do workplace demands, work resources, and time-varying factors relate to educator stress, and how much uncertainty remains after accounting for repeated observations within educators?

## Setup

```{r setup}
source("R/data_dictionary.R")
source("R/prepare_data.R")
source("R/model_functions.R")

library(dplyr)
library(ggplot2)
```

## Raw Data Inventory

```{r raw-inventory}
raw_inventory <- raw_file_inventory("data/raw")
raw_inventory
```

## Time Point Files

```{r timepoint-summary}
timepoint_data <- load_timepoint_data("data/raw")
timepoint_summary <- summarise_timepoints(timepoint_data)
timepoint_summary
```

## Column Inventory

```{r column-inventory}
column_inventory <- csv_column_inventory("data/raw")

column_inventory |>
  count(file, name = "columns") |>
  arrange(file)
```

## Candidate Psychological Constructs

This first pass searches column names for likely stress, demand, resource, well-being, and educator identifiers. These candidates will be verified against the codebook before modeling.

```{r candidate-columns}
candidate_patterns <- c(
  "stress", "burn", "well", "demand", "resource", "support",
  "depress", "anx", "effic", "satisf", "job", "work", "teacher",
  "center", "site", "id"
)

candidates <- find_candidate_columns(column_inventory, candidate_patterns)

candidates |>
  arrange(file, column) |>
  select(file, column, missing_rate) |>
  head(80)
```

## Next Modeling Plan

The next step is to use the codebook to define:

- the primary stress or well-being outcome;
- educator identifier;
- center or site identifier, if available;
- repeated-measure time variable;
- job demand predictors;
- job resource predictors;
- demographic or professional covariates.

The first Bayesian model will be intentionally simple. Once variable definitions are verified, the project will move to a hierarchical longitudinal model.
