---
title: "Bayesian Hierarchical Modeling of Workplace Stress"
author: "Washington Casamen Nolasco"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    highlight-style: github
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
knitr:
  opts_knit:
    root.dir: "../"
---

## Overview

This report develops a Bayesian hierarchical analysis of workplace stress and well-being among early childhood educators. The project focuses on repeated measurements, uncertainty, and practical interpretation for behavioral data science and occupational psychology.

## Research Question

How do workplace demands, work resources, and time-varying factors relate to educator stress, and how much uncertainty remains after accounting for repeated observations within educators?

## Setup

```{r setup}
source("R/data_dictionary.R")
source("R/prepare_data.R")
source("R/model_functions.R")

library(dplyr)
library(ggplot2)
```

## Raw Data Inventory

```{r raw-inventory}
raw_inventory <- raw_file_inventory("data/raw")
raw_inventory
```

## Time Point Files

```{r timepoint-summary}
timepoint_data <- load_timepoint_data("data/raw")
timepoint_summary <- summarise_timepoints(timepoint_data)
timepoint_summary
```

## Column Inventory

```{r column-inventory}
column_inventory <- csv_column_inventory("data/raw")

column_inventory |>
  count(file, name = "columns") |>
  arrange(file)
```

## Analysis 1 - Longitudinal Structure

Before choosing a Bayesian model, we need to verify whether the same educators are observed across time points. This determines whether the model can use repeated-measure partial pooling, for example `(1 | StudyID)`, or whether the analysis should be treated as mostly cross-sectional by wave.

```{r analysis-1-id-audit}
id_audit <- audit_study_ids(timepoint_data, id_col = "StudyID")
id_audit
```

```{r analysis-1-id-overlap}
id_overlap <- build_id_overlap_matrix(timepoint_data, id_col = "StudyID")
id_overlap
```

**Interpretation guide:** The diagonal shows the number of unique educators in each wave. Off-diagonal cells show how many educators appear in both waves. If overlap is high, a longitudinal hierarchical model is appropriate. If overlap is low, we may need a simpler repeated-wave or available-case approach.

## Analysis 2 - Candidate Scale Scores

The raw files contain many item-level survey columns. For a first Bayesian model, we should prefer validated or precomputed scale scores when available, then verify their meaning in the codebook. This chunk searches for columns that look like totals, sums, means, averages, or scale scores.

```{r analysis-2-summary-columns}
summary_columns <- find_summary_columns(timepoint_data)

summary_columns |>
  arrange(time_point, column)
```

```{r analysis-2-scale-profiles}
scale_profiles <- profile_numeric_columns(timepoint_data, summary_columns)

scale_profiles |>
  arrange(time_point, column)
```

**Interpretation guide:** These columns are candidates, not final constructs. We will use the codebook to decide which score should become the primary outcome, such as perceived stress, burnout, depressive symptoms, job satisfaction, or teacher stress.

## Candidate Psychological Constructs

This first pass searches column names for likely stress, demand, resource, well-being, and educator identifiers. These candidates will be verified against the codebook before modeling.

```{r candidate-columns}
candidate_patterns <- c(
  "stress", "burn", "well", "demand", "resource", "support",
  "depress", "anx", "effic", "satisf", "job", "work", "teacher",
  "center", "site", "id"
)

candidates <- find_candidate_columns(column_inventory, candidate_patterns)

candidates |>
  arrange(file, column) |>
  select(file, column, missing_rate) |>
  head(80)
```

## Next Modeling Plan

The next step is to use the codebook to define:

- the primary stress or well-being outcome;
- educator identifier;
- center or site identifier, if available;
- repeated-measure time variable;
- job demand predictors;
- job resource predictors;
- demographic or professional covariates.

The first Bayesian model will be intentionally simple. Once variable definitions are verified, the project will move to a hierarchical longitudinal model.
